\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-04-24 forwards-and-backwards.tex
up to section 7

\comment{get equations to restart numbering in sections}

@@

We introduce a broad class of mechanistic spatial models to describe how spatially 
heterogeneous populations live, die, and reproduce.
Individuals are represented by points of a point measure,
whose birth and death rates can depend 
both on spatial position and local population density, defined at a location
to be the convolution of the point measure
with a suitable non-negative integrable kernel
centred on that location. We pass to three different scaling limits:
an interacting superprocess, a nonlocal partial differential equation (PDE), 
and a classical PDE. 
The classical PDE is obtained both by a two-step convergence argument, in which we 
first scale time and population size and
pass to the nonlocal PDE, 
and then scale the kernel that 
determines local population density; and 
in the important special case in which the limit is a 
reaction diffusion equation, directly 
by simultaneously scaling the kernel and the
timescale and population size in our individual based model.

A novelty of our model is that we explicitly
model a juvenile phase: the number of juveniles produced by an individual depends
on local population density at the location of the parent. 
Juvenile offspring are 
thrown off in a (possibly heterogeneous, anisotropic) Gaussian distribution 
around the location of the parent; they then reach (instant) maturity with a probability
that can depend on the local population density
% (possibly measured through convolution with a different kernel from that used in determining fecundity) 
at the location at which they land. 
Although we only record mature individuals, a trace of this two-step description 
remains in our population models, resulting in novel limits in which the spatial
motion is governed by a nonlinear diffusion.

@@

As one takes a journey, long or short, the landscape changes:
forests thicken or thin or change their composition;
springtime grasslands host intergrading mosaics of different types of flowers.
The aim of this paper is to introduce and study
a broad class of mechanistic spatial models
that might describe how spatially heterogeneous populations live, die, and reproduce.
Questions that we (start to) address include:
How does population density change across space and time?
How might we learn about the underlying dynamics from genealogical or genetic data?
And, how does genetic ancestry spread across geography
when looking back through time in these populations?

@@

Most models do not distinguish between juveniles and adults, so,
for example, the number of adults produced by a single parent is determined
by the degree of crowding at the location of the parent.
Although we shall similarly only follow the adult population, in formulating the dynamics of the
models we shall distinguish
between production of juveniles, which will depend upon the location of 
the adult, and their successful establishment, which will depend on the
location in which a juvenile lands. The result is that not only the absolute 
number, but also the spatial distribution
around their parent, 
of those offspring that survive to adulthood
will depend upon the local population 
density.

@@

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we expect that if we take a second limit of $\epsilon \to 0$,
thus scaling the kernels appearing in $r$, $\gamma$, and $F$ and making interactions pointwise,
we should recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which we 
retain a signal of our two stage approach to modelling in the nonlinear diffusion term
and the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.
Equations of this form have been studied extensively
in the context of spatial ecology;
see for instance \citet{lam2023introduction} and \citet{cantrell2004spatial}.

@@

We remark that such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. 
\comment{FROM ALISON: I rewrote the rest of this paragraph. Missing references
are in the latex file, next to the citation, but commented out}
Following the now classical contributions 
of~\citet{demasi/ferrari/lebowitz:1986, demasi/presutti:1991,
oeschlaeger:1985},
%@article{demasi/ferrari/lebowitz:1986,
%  title={Reaction-diffusion equations for interacting particle system},
%  author={De Masi, A. and Ferrari, P. A. and Lebowitz, J. L.},
%  journal={J. Stat. Phys.},
%  year={1986},
%  volume={44},
%  pages={589-644},
%  DOI = {10.1007/BF01011311},
%}
% A. De Masi and E. Presutti. Mathematical methods for hydrodynamic limits,
%volume 1501 of Lecture Notes in Mathematics. Springer-Verlag, Berlin, 1991.
%
% K. Oeschl\"ager. A law of large numbers for moderately interacting 
% diffusion processes. Z. Warsch. Verw. Gebiete. 69(2):279-322, 1985
much of this work has focused on lattice based models with one particle
per site, or on systems with a fixed number, $N$,
of interacting diffusions as $N\to\infty$.
For systems of proliferating particles,
as considered for example by 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019,
flandoli/huang:2021}, 
% K. Oeschl\"ager. On the derivation of reaction-diffusion equations as limits
% of systems of moderately interacting stochastic processes. Probab. Th. Rel.
% Fields 82, 565--586, (1989)
%
% F. Flandoli, M. Leimbach, and C. Olivera. Uniform convergence of 
%proliferating particles to the FKPP equation. J. Math. Anal. Appl. 473(1):
%27--52, (2019)
an additional challenge (also apparent in our models), is 
the control of concentration of particles. 
We follow 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019} in considering
`moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas 
\citet{flandoli/huang:2021} also consider the situation in which
that number remains finite. 
We refer to \citet{flandoli/huang:2021} for a more thorough 
literature review, 
but note that both our model and scaling differ from those considered
in the body of work discussed there: whereas in those settings, the only 
scalings are the number of particles in the system and the size of the
neighbourhood over which individuals interact with one another, in
keeping with the vast literature on continuous state branching models, we 
also scale time and so must ensure that births are 
adequately compensated by deaths to prevent the population from exploding.

The history of a natural population is often only accessible indirectly, 
through patterns of genetic diversity that have been laid down; from 
genetic data, one can try to infer the genealogical trees that relate 
individuals in a sample from the population, and these have been shaped 
by its history \citep{neigel1993application,kelleher2019inferring}.
It is therefore of interest to establish information about the distribution of genealogical trees 
under our population model.
To achieve this we exploit a lookdown construction. 
Lookdown constructions were first introduced in~\cite{donnelly/kurtz:1996} to provide 
a mechanism for retaining information about genealogical relationships between individuals
sampled from a population evolving according to the Moran model when passing to the 
infinite population limit.
Since then, they have been extended to a huge range of models. Of particular
relevance to our work here are the papers of~\cite{kurtz/rodrigues:2011} 
and~\cite{etheridge/kurtz:2019}, in which
lookdown constructions are provided for a wide variety of population models, including
spatially structured branching processes.

In general, even armed with a lookdown construction, calculation of relevant statistics 
of the genealogy remains a difficult question. 
However, in special
circumstances, some progress can be made. As an illustration, 
we shall consider a scenario that
has received a great deal of attention in recent years, in which 
a population is expanding into new territory as a travelling wave. 
In Section~\ref{ancestral lineages for travelling waves}
we shall consider three different (deterministic) wavefronts, moving 
across $\IR^1$, and we shall explore the position of a single ancestral 
lineage relative to that front.

Most work on this topic has focused on models that caricature the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. 
For instance, \citep{ghosh2022emergent}
derives a stochastic version of~\eqref{semilinear PDE} to describe abundances of mutant bacteria strains
along the human gut.
For very large $N$, and on suitable timescales, 
starting with the pioneering work of~\cite{brunet/derrida/mueller/munier:2006},
a considerable body of evidence has been amassed to underpin the
conjecture that for this, and a wide class of 
related models, 
genealogies converge to a 
Bolthausen-Sznitman coalescent. %, e.g.~\cite{brunet/derrida/mueller/munier:2007}.
This reflects the fact that, for this equation, 
ancestral lineages become trapped in 
the wavefront, where the growth rate of the population is highest. 
Once there, they will experience 
rapid periods of coalescence 
corresponding to significant proportions of individuals in the front being 
descended from particularly reproductively successful ancestors.

If one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
\begin{equation}
	\label{Birzu nonlinearity}
	wF(w)=w(1-w)(Cw-1),
\end{equation}
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from ``pulled'' to ``pushed'', 
\citep{birzu/hallatschek/korolev:2017, birzu/hallatschek/korolev:2021}.
\comment{The missing reference is
 G. Birzu, O. Hallatshcek and K. Korolev. Genealogical structure
 changes as range expansions transition from pushed to pulled. 
 PNAS 118(34) (2021) 
 https://doi.org/10.1073/pnas.2026746118}
In that setting, the genealogies will be quite different
from the Fisher-KPP case. 
%\comment{Could refer to more recent Birzu et al paper here.}
For example, \citet{etheridge/penington:2022}
show that for a discrete space model corresponding to this 
nonlinearity with $C>2$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet
it is typically at a location in which population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

The shape of the wave is not determined solely by the reaction term.
For example, as a result of the nonlinear diffusion, 
for suitable initial conditions, the solution to the one-dimensional 
porous medium equation with logistic growth~(\ref{PME1}) converges to a travelling
wave with a sharp cut-off; i.e., in contrast to the classical 
Fisher KPP equation, the solution at time $t$
vanishes beyond $x=x_0+ct$ for some constant wavespeed $c>0$, \citep{kamin/rosenau:2004}.
As a first step towards understanding what we should expect in models with
nonlinear diffusion, one can ask about the position of an ancestral lineage
relative to the wavefront in the deterministic models. In 
Section~\ref{ancestral lineages for travelling waves}
we shall see that in our framework, even with logistic growth, the nonlinear diffusion 
corresponding to the porous medium equation results
in a stationary distribution for the ancestral lineage
that is concentrated behind the wavefront, leading 
us to conjecture that in the stochastic equation
the cooperative behaviour captured by the nonlinear diffusion will also 
result in a qualitatively different pattern of coalescence to that seen under the 
stochastic Fisher-KPP equation. Indeed, we believe that it should be feasible to 
show that in an appropriate limit one recovers a Kingman coalescent.

@@

\comment{TODO: This paragraph to be rewritten last PETER}

In this paper we study scaling limits of spatial population models,
obtaining convergence of both the population process
(i.e., the population density as a function of time,
although strictly speaking it is a measure that may not have a density)
and of lineages traced back through such a population.
We retain information about lineages as we pass to the scaling limit
by means of a lookdown construction.
First, in Section~\ref{sec: Model and main results},
we describe the model and the main results
(Theorems~\ref{thm:nonlocal_convergence}, \ref{thm:local_convergence}, and~\ref{thm:lineages}).
Next, in Section~\ref{sec:applications}, we discuss a few striking consequences of these results,
namely, X Y and Z \comment{TODO}.
In Section~\ref{sec:heuristics}, we provide heuristic explanations
of why the theorems ought to be true,
and some key ideas behind them.
In Section \ref{sec:lookdown} we define and discuss the lookdown construction.
This is followed by the technical proofs in Section~\ref{sec:proofs}.

\comment{What's in the Appendix?}

@@

Population dynamics are controlled by three quantities,
birth ($\gamma$), establishment ($r$), and death ($\mu$),
each of which can depend on spatial location and local population density
in a way specified below.
Each individual gives birth at rate $\gamma$ to a single (juvenile) offspring,
which is dispersed according to a kernel $q(x, \cdot)$ away from the location $x$ of the parent.
We assume that $q$ is the density of a multivariate Gaussian,
allowing a nonzero mean and anisotropic variance.
Both the mean and covariance can change across space,
but do not depend on population density.
The offspring does not necessarily survive to be counted in the population:
it ``establishes'' with probability $r$,
or else it dies immediately.
Independently, each individual dies with rate $\mu$.

We aim to capture universal behaviour by passing to a scaling limit. Specifically, 
we shall take the ``density'', $N$, to infinity,
and also scale time by a factor of $\theta = \theta(N)$,
in such a way that defining $\eta^N(t) = X(\theta t) / N$,
the process $\{\eta^N(t)\}_{t \ge 0}$
will converge to a suitable measure-valued process
as $N$ and $\theta$ tend to infinity,
with the nature of the limit depending on how they tend to infinity together.
Evidently, we shall also need to scale the dispersal kernel if we are to obtain a
nontrivial limit. We write $q_\theta(x,\cdot)$ for the density of the 
multivariate Gaussian obtained by
multiplying the mean and variance components corresponding to $q(x,\cdot)$ by $1/\theta$.

@@

We shall write $\mu_\theta(x, X/N)$ for the per-capita death rate of mature individuals
in the population.
In order for the 
population density to change over timescales of order $\theta$, we should like the net 
per capita reproductive rate 
to scale as $1/\theta$. 
In classical models, in which $r$, $\gamma$, and $\mu$ are constant, 
this quantity is simply $r\gamma -\mu$. 
Here, because production of juveniles and their establishment are
mediated by population density measured relative to different points, 
the net reproductive rate will take a more complicated form. 
In particular, the total rate of production of mature offspring by an individual at $x$ will be
\begin{equation} \label{eqn:total_repro}
	\gamma\big(x,\smooth{\gamma}X(x)/N\big)\int r\big(y, \smooth{r}X(y)/N\big)q_{\theta}(x,dy).
\end{equation}
Nonetheless, it will be convenient to define the death rate $\mu_\theta$ in terms of 
its deviation from $r\gamma$. 
To this end, we define
the death rate of an individual at $x$ to be
\begin{align} \label{eqn:mu_defn}
    \mu_\theta(x)
    =
    \max\left\{0, 
        r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N)
    \right\} ,
\end{align}
where this equation defines $F(x, m) : \IR^d \times \IR_{\ge 0} \to \IR$,
and $\rho_F$ is again a smooth kernel. 
The function $F$ is nearly
the net per capita reproductive rate, scaled by $\theta$,
and would be equal to it in a nonspatial model;
but, as can be seen from \eqref{eqn:total_repro},
differs because an offspring's establishment probability is measured at their new location
rather than that of their parent.
For the most part, we work with $F$ instead of $\mu$.

Note that each of the three demographic parameters $r$, $\gamma$, $F$,
depends on local density,
measured by convolution with a smooth kernel. Each of the three kernels can be different.
As a result, death rate depends (in principle) on population densities measured in 
three different ways, 
so that we could write 
$\mu_\theta(x) = \mu_\theta(x, \smooth{\gamma} X(x) / N, \smooth{r} X(x) / N, \smooth{F} X(x) / N)$.
This may seem unnecessarily complex.
However, not only is it natural from a 
biological perspective, it also turns out to be necessary if we are
to capture nontrivial examples in our scaling results.

@@

For each $N$ and $\theta$, the rescaled process
$\big(\eta^N_t\big)_{t\geq 0} := \big(X(\theta t)/N\big)_{t\geq 0}$, 
takes values in the space of c\`adl\`ag paths in
$\measures$ (the space of finite measures on $\IR^d$
endowed with the weak topology). In fact $\eta_t^N$ will be a purely atomic
measure comprised of atoms of mass $1/N$.

@@

In summary, at each time $t$, $\eta^N_t$ is purely atomic, consisting of atoms of mass $1/N$.
At instantaneous rate $\theta \gamma(x, \eta^N_t) N \eta^N_t(dx)$ an offspring of mass $1/N$ is 
produced at location $x$; it disperses to 
a location $y$ offset from $x$
by an independent Gaussian random variable with mean $\meanq(x) / \theta$
and covariance matrix $\covq(x) / \theta$, and once there establishes instantaneously
with probability $r(y, \eta^N_t)$, or else dies. 
At instantaneous rate $\theta \mu_\theta(x, \eta^N_t) N \eta^N_t(dx)$ an individual at location 
$x$ dies.
Note that the process 
$\left(\eta^N_t\right)_{t\ge 0}$, which
records numbers and locations of adult individuals, is just a scaled spatial birth and 
death process. If, for example, we insist that $\gamma(x,m)$ is bounded, then 
existence (and in particular non-explosion) is guaranteed by comparison with
a pure birth process. We do not dwell on this, as we shall require more stringent conditions
if we are to pass to the limit as $\theta$ and $N$ tend to infinity.

@@

Our main results depend on two dichotomies:
Is the limiting process deterministic or a (generalized) superprocess?
And, are interactions pointwise in the limit or nonlocal?
Below we have results for deterministic limits with pointwise and nonlocal interactions,
and for superprocess limits with nonlocal interactions.
See Figure \ref{fig:super_vs_det_2d} for snapshots of the population
from direct simulation of the process using SLiM \citep{haller_slim_2019}.

@@

We now turn to some examples and applications.

>>>>

\comment{TODO: ALISON SAYS
The proof of Lemma 6.7 is now problematic. We'll be hard pushed to find a sequence of functions increasing to x, or indeed -x. Easily fixed of course, but I think that we should anyway change the condition of the theorem to be a bit more general, so say that we assume that $\langle f(x),\eta^N_0(dx)\rangle$  is uniformly bounded for some non-negative, twice differentiable with uniformly bounded first and second derivatives, function $f$ with $f(x)\to\infty$ as $x\to\infty$ and then use Lemma 6.1 and Gronwall to deduce $\langle f,\eta_t\rangle$ has bounded mean for any limit point and so Markov's inequality tells us no mass escaped to infinity.
}

\comment{``reaction diffusion equation'' or ``reaction-diffusion equation''?}

@@

We introduce a broad class of mechanistic spatial models to describe how spatially 
heterogeneous populations live, die, and reproduce.
Individuals are represented by points of a point measure,
whose birth and death rates can depend 
both on spatial position and local population density, defined at a location
to be the convolution of the point measure
with a suitable non-negative integrable kernel
centred on that location. We pass to three different scaling limits:
an interacting superprocess, a nonlocal partial differential equation (PDE), 
and a classical PDE. 
The classical PDE is obtained both by a two-step convergence argument, in which we 
first scale time and population size and
pass to the nonlocal PDE, 
and then scale the kernel that 
determines local population density; and 
in the important special case in which the limit is a 
reaction diffusion equation, directly 
by simultaneously scaling the kernel width,
timescale and population size in our individual based model.

A novelty of our model is that we explicitly model a juvenile phase.
The number of juveniles produced by an individual depends
on local population density at the location of the parent;
these juvenile offspring are 
thrown off in a (possibly heterogeneous, anisotropic) Gaussian distribution 
around the location of the parent; they then reach (instant) maturity with a probability
that can depend on the local population density
% (possibly measured through convolution with a different kernel from that used in determining fecundity) 
at the location at which they land. 
Although we only record mature individuals, a trace of this two-step description 
remains in our population models, resulting in novel limits in which the spatial
dynamics are governed by a nonlinear diffusion.

@@

As one takes a journey, long or short, the landscape changes:
forests thicken or thin or change their composition;
even in flat plains
springtime grasslands host intergrading mosaics of different types of flowers.
The aim of this paper is to introduce and study
a broad class of mechanistic spatial models
that might describe how spatially heterogeneous populations live, die, and reproduce.
Questions that we (start to) address include:
How does population density change across space and time?
How might we learn about the underlying dynamics from genealogical or genetic data?
And, how does genetic ancestry spread across geography
when looking back through time in these populations?

@@

Often models do not distinguish between juveniles and adults, so,
for example, the number of adults produced by a single parent is determined only
by the degree of crowding at the location of the parent.
Although we shall similarly only follow the adult population, in formulating the dynamics of the
models we shall distinguish
between production of juveniles, which will depend upon the location of 
the adult, and their successful establishment, which will depend on the
location in which a juvenile lands. The result is that not only the absolute 
number, but also the spatial distribution
around their parent, 
of those offspring that survive to adulthood
will depend upon the local population 
density.

@@

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we expect that if we take a second limit of $\epsilon \to 0$,
thus scaling the kernels appearing in $r$, $\gamma$, and $F$ and making interactions pointwise,
we should recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which
the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.
Equations of this form have been studied extensively
in the context of spatial ecology;
see for instance \citet{lam2023introduction} and \citet{cantrell2004spatial},
and in many other fields;
for instance, \citet{ghosh2022emergent}
derive a stochastic version of~\eqref{semilinear PDE} to describe abundances of mutant bacteria strains
along the human gut,
while \citet{li/buenzli/simpson:2022} study the effects of nonlinear diffusion
on long-term survival of a lattice-based interacting particle system.
% MORE DETAILS:
% \cite{li/buenzli/simpson:2022} write down an interacting particle
% system on a hexagonal lattice (with at most one individual per site) that incorporates an effect
% of local crowding on individual motion. Their equation
% corresponds to a reaction-diffusion equation with nonlinear diffusion and a 
% bistable reaction term of the form~(\ref{Birzu nonlinearity}) and they
% investigate the interplay between nonlinear diffusion and long-term 
% survival or extinction of the population.

@@

Such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. 
Following the now classical contributions 
of~\citet{demasi/ferrari/lebowitz:1986, demasi/presutti:1991, oeschlaeger:1985},
much of this work has focused on lattice based models with one particle
per site, or on systems with a fixed number, $N$,
of interacting diffusions as $N\to\infty$.
For systems of proliferating particles,
as considered for example by 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019, flandoli/huang:2021}, 
an additional challenge (also apparent in our models), is 
the control of concentration of particles. 
We follow 
\citet{oelschlaeger:1989, flandoli/leimbach/olivera:2019} in considering
`moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas 
\citet{flandoli/huang:2021} also consider the situation in which
that number remains finite. 
We refer to \citet{flandoli/huang:2021} for a more thorough 
literature review, 
but note that both our model and scaling differ from those considered
in the body of work discussed there: whereas in those settings, the only 
scalings are the number of particles in the system and the size of the
neighbourhood over which individuals interact with one another, in
keeping with the vast literature on continuous state branching models, we 
also scale time and so must ensure that births are 
adequately compensated by deaths to prevent the population from exploding.

The history of a natural population is often only accessible indirectly, 
through patterns of genetic diversity that have been laid down; from 
genetic data, one can try to infer the genealogical trees that relate 
individuals in a sample from the population, and these have been shaped 
by its history \citep[see e.g.,][]{neigel1993application,kelleher2019inferring}.
It is therefore of interest to establish information about the distribution of genealogical trees 
under our population model, which we do with a lookdown construction. 
Lookdown constructions were first introduced in~\cite{donnelly/kurtz:1996} to provide 
a mechanism for retaining information about genealogical relationships between individuals
sampled from a population evolving according to the Moran model when passing to the 
infinite population limit.
Since then, they have been extended to a wide range of models. Of particular
relevance to our work here are the papers of~\cite{kurtz/rodrigues:2011} 
and~\cite{etheridge/kurtz:2019}, in which
lookdown constructions are provided for a wide variety of population models, including
spatially structured branching processes.

In general, even armed with a lookdown construction, calculation of relevant statistics 
of the genealogy remains a difficult question. 
However, in special
circumstances, some progress can be made. As an illustration, 
we shall consider a scenario that
has received a great deal of attention in recent years, in which 
a population is expanding into new territory as a travelling wave. 
In Section~\ref{ancestral lineages for travelling waves}
we shall describe the motion of a single ancestral lineage relative to
three different (deterministic) wavefronts across $\IR^1$.

Most work on the topic of ``waves'' of expanding populations
has focused on models that caricature the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. 
Starting with the pioneering work of~\cite{brunet/derrida/mueller/munier:2006},
a considerable body of evidence has been amassed to underpin the
conjecture that for this, and a wide class of related models, 
genealogies converge on suitable timescales in the infinite density limit
to a Bolthausen-Sznitman coalescent. %, e.g.~\cite{brunet/derrida/mueller/munier:2007}.
This reflects the fact that, for this equation, 
ancestral lineages become trapped in 
the wavefront, where the growth rate of the population is highest. 
Once there, they will experience 
rapid periods of coalescence 
corresponding to significant proportions of individuals in the front being 
descended from particularly reproductively successful ancestors.

If one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
\begin{equation}
	\label{Birzu nonlinearity}
	wF(w)=w(1-w)(Cw-1),
\end{equation}
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from ``pulled'' to ``pushed'', 
\citep{birzu/hallatschek/korolev:2017, birzu/hallatschek/korolev:2021},
and so the genealogies will be quite different from the Fisher-KPP case. 
For example, \citet{etheridge/penington:2022}
show that for a discrete space model corresponding to this 
nonlinearity with $C>2$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet
it is typically at a location in which population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

The shape of the wave is not determined solely by the reaction term.
For example, as a result of the nonlinear diffusion, 
for suitable initial conditions, the solution to the one-dimensional 
porous medium equation with logistic growth~(\ref{PME1}) converges to a travelling
wave with a sharp cut-off; i.e., in contrast to the classical 
Fisher KPP equation, the solution at time $t$
vanishes beyond $x=x_0+ct$ for some constant wavespeed $c>0$ \citep{kamin/rosenau:2004}.
As a first step towards understanding what we should expect in models with
nonlinear diffusion, one can ask about the position of an ancestral lineage
relative to the wavefront in the deterministic models. In 
Section~\ref{ancestral lineages for travelling waves}
we shall see that in our framework, even with logistic growth, the nonlinear diffusion 
corresponding to the porous medium equation results
in a stationary distribution for the ancestral lineage
that is concentrated behind the wavefront, leading 
us to conjecture that in the stochastic equation
the cooperative behaviour captured by the nonlinear diffusion will also 
result in a qualitatively different pattern of coalescence to that seen under the 
stochastic Fisher-KPP equation. Indeed, we believe that it should be feasible to 
show that in an appropriate limit one recovers a Kingman coalescent.

@@

In this paper we study scaling limits of spatial population models,
obtaining convergence of both the population process
(i.e., the population density as a function of time,
although strictly speaking it is a measure that may not have a density)
and of lineages traced back through such a population.
We retain information about lineages as we pass to the scaling limit
by means of a lookdown construction.

In what follows we first study various scaling limits of the spatial population process,
and then turn our attention to lineages traced back through these populations.
First, in Section~\ref{sec: Model and main results},
we describe the model and the main results,
Theorems~\ref{thm:nonlocal_convergence}, \ref{thm:local_convergence}, and~\ref{thm:lineages}.
Next, in Section~\ref{sec:applications}, we discuss a few striking consequences of these results
regarding the behavior of genealogies in traveling waves,
the appearance of periodic ``clumps'' in seemingly homogeneous population models,
and identifiability of the underlying dynamics from a stationary population profile.
In Section~\ref{sec:heuristics}, we provide heuristic explanations
of why the theorems ought to be true,
and some key ideas behind them,
and in Section~\ref{sec:lookdown} we define and discuss the lookdown construction.
Proofs of the results begin in Section~\ref{sec:proofs},
which proves results for population models with nonlocal interactions,
while Section~\ref{subsec:one step convergence proof} gives the more difficult proof
for the case when interaction distances also go to zero in the limit.
Finally, Section~\ref{sec:lookdown_proofs} gives proofs for convergence of the lookdown process
and the associated results for the motion of lineages.
The Appendix contains a few more techinical and less central lemmas.
The results are illustrated in a few places with individual-based simulations,
made using SLiM \citep{haller_slim_2019},
but these are provided for visualization and we do not embark on numerical study.

@@

Population dynamics are controlled by three quantities,
birth ($\gamma$), establishment ($r$), and death ($\mu$),
each of which can depend on spatial location and local population density
in a way specified below.
Each individual gives birth at rate $\gamma$ to a single (juvenile) offspring,
which is dispersed according to a kernel $q(x, \cdot)$ away from the location $x$ of the parent.
We assume that $q$ is the density of a multivariate Gaussian,
allowing a nonzero mean and anisotropic variance.
Both the mean and covariance of $q$ can change across space,
but do not depend on population density.
The offspring does not necessarily survive to be counted in the population:
it ``establishes'' with probability $r$,
or else it dies immediately.
Independently, each individual dies with rate $\mu$.

We aim to capture universal behaviour by passing to a scaling limit. Specifically, 
we shall take the ``density'', $N$, to infinity,
and also scale time by a factor of $\theta = \theta(N)$,
in such a way that defining $\eta^N(t) = X(\theta t) / N$,
the process $\{\eta^N(t)\}_{t \ge 0}$
will converge to a suitable measure-valued process
as $N$ and $\theta$ tend to infinity,
with the nature of the limit depending on how they tend to infinity together.
Evidently, we also need to scale the dispersal kernel if we are to obtain a
nontrivial limit, for which we use $q_\theta(x,\cdot)$,
the density of the multivariate Gaussian obtained by
multiplying the mean and variance components of $q(x,\cdot)$ by $1/\theta$.

@@

We shall write $\mu_\theta(x, X/N)$ for the per-capita death rate of mature individuals
in the population.
In order for the 
population density to change over timescales of order $\theta$, we should like the net 
per capita reproductive rate 
to scale as $1/\theta$. 
In classical models, in which $r$, $\gamma$, and $\mu$ are constant, 
this quantity is simply $r\gamma -\mu$. 
Here, because production of juveniles and their establishment are
mediated by population density measured relative to different points, 
the net reproductive rate will take a more complicated form. 
In particular, the total rate of production of mature offspring by an individual at $x$ will be
\begin{equation} \label{eqn:total_repro}
	\gamma\big(x,\smooth{\gamma}X(x)/N\big)\int r\big(y, \smooth{r}X(y)/N\big)q_{\theta}(x,dy).
\end{equation}
Nonetheless, it will be convenient to define the death rate $\mu_\theta$ in terms of 
its deviation from $r\gamma$. 
To this end, we define
the death rate of an individual at $x$ to be
\begin{align} \label{eqn:mu_defn}
    \mu_\theta(x)
    =
    \max\left\{0, 
        r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N)
    \right\} ,
\end{align}
where this equation defines $F(x, m) : \IR^d \times \IR_{\ge 0} \to \IR$,
and $\rho_F$ is again a smooth kernel. 
The function $F$ is nearly
the net per capita reproductive rate, scaled by $\theta$,
and would be equal to it in a nonspatial model;
but, as can be seen from \eqref{eqn:total_repro},
differs because an offspring's establishment probability is measured at their new location
rather than that of their parent.
For the most part, we work with $F$ instead of $\mu_\theta$.

So, each of the three demographic parameters $r$, $\gamma$, and $F$,
depends on local density, measured by convolution with a smooth kernel,
each of which can be different.
As a result, death rate depends (in principle) on population densities measured in 
three different ways, 
so that we could write 
$\mu_\theta(x) = \mu_\theta(x, \smooth{\gamma} X(x) / N, \smooth{r} X(x) / N, \smooth{F} X(x) / N)$.
This may seem unnecessarily complex.
However, not only is it natural from a 
biological perspective, it also turns out to be convenient for
capturing nontrivial examples in the scaling limit.

@@

For each $N$ and $\theta$, we study primarily the process
with mass scaled by $N$ and time scaled by $\theta$,
$$ \big(\eta^N_t\big)_{t\geq 0} := \big(X(\theta t)/N\big)_{t\geq 0} , $$ 
which takes values in the space of c\`adl\`ag paths in
$\measures$ (the space of finite measures on $\IR^d$
endowed with the weak topology). In fact $\eta_t^N$ will be a purely atomic
measure comprised of atoms of mass $1/N$.

@@

In summary, at each time $t$, $\eta^N_t$ is purely atomic, consisting of atoms of mass $1/N$
(which are the individuals).
At instantaneous rate $\theta \gamma(x, \eta^N_t) N \eta^N_t(dx)$ an offspring of mass $1/N$ is 
produced at location $x$; it disperses to 
a location $y$ offset from $x$
by an independent Gaussian random variable with mean $\meanq(x) / \theta$
and covariance matrix $\covq(x) / \theta$, and once there establishes instantaneously
with probability $r(y, \eta^N_t)$, or else dies. 
At instantaneous rate $\theta \mu_\theta(x, \eta^N_t) N \eta^N_t(dx)$ an individual at location 
$x$ dies.
Note that the process 
$\left(\eta^N_t\right)_{t\ge 0}$, which
records numbers and locations of adult individuals, is just a scaled spatial birth and 
death process. If, for example, we insist that $\gamma(x,m)$ is bounded, then 
existence (and in particular non-explosion) is guaranteed by comparison with
a pure birth process. We do not dwell on this, as we shall require more stringent conditions
if we are to pass to the limit as $\theta$ and $N$ tend to infinity.

@@

Our main results depend on two dichotomies:
Is the limiting process deterministic or a (generalized) superprocess?
And, are interactions pointwise in the limit or nonlocal?
See Figure \ref{fig:super_vs_det_2d} for snapshots of the population
from direct simulation of the process using SLiM \citep{haller_slim_2019}
illustrating this first dichotomy.
Below we have results for deterministic limits with pointwise and nonlocal interactions,
and for superprocess limits with nonlocal interactions.

@@

We now discuss some consequences of these results.
